---
title: "PhyloseqDEJUCloaca"
author: "Tricia"
date: "`r Sys.Date()`"
output: pdf_document
---

##Load required packages
```{r warning=FALSE, message=FALSE}
library(ggplot2)
library(car)
library(dada2)
library(phyloseq)
library(vegan)
library(nlme)
library(plyr)
library(Biostrings)
library(tidyverse)
library(ggpubr)
library(decontam) #isContaminant
library(BiMiCo) #rm_nonbac
library(microbiome)
library(MicEco)
library(lme4)
library(dplyr)
```

##Set seed for reproducibility
```{r}
set.seed(777)
```

##Load taxa and seqtab files to start here
```{r}
load("../RData/taxa.RData")
load("../RData/seqtab.nochim.RData")
```

##import metadata
```{r}
metadata<-read.csv("../metadata.csv", header=TRUE, row.names = 1)
```

##Create phyloseq object
```{r}
#make sure the seqtab.nochim and taxa objects are loaded
physeq <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), 
               sample_data(metadata), 
               tax_table(taxa))
physeq
```
##inspect library sizes
```{r Inspect Library Size}
df <- as.data.frame(sample_data(physeq)) # Put sample_data into a ggplot-friendly data.frame
df$LibrarySize <- sample_sums(physeq)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
ggplot(data=df, aes(x=Index, y=LibrarySize, color=control)) + geom_point()
```
##identify contaminants
```{r Identify Contaminants at Higher Freq in Negative Controls}
sample_data(physeq)$is.neg <- sample_data(physeq)$control == "yes"
contamdf.prev <- isContaminant(physeq, method="prevalence", neg="is.neg", threshold=0.5) #identify contaminants in higher frequency in negative controls
table(contamdf.prev$contaminant)
head(which(contamdf.prev$contaminant))
```

##remove control samples
```{r Remove Control Samples}
# Make phyloseq object of presence-absence in negative controls and true samples
physeq.pa <- transform_sample_counts(physeq, function(abund) 1*(abund>0))
physeq.pa.neg <- prune_samples(sample_data(physeq.pa)$control == "yes", physeq.pa)
physeq <- prune_samples(sample_data(physeq.pa)$control == "no", physeq.pa) #this will contain positives
physeq
```
##graph controls
```{r Graph Controls}
df.pa <- data.frame(pa.pos=taxa_sums(physeq), pa.neg=taxa_sums(physeq.pa.neg),
                      contaminant=contamdf.prev$contaminant)
ggplot(data=df.pa, aes(x=pa.neg, y=pa.pos, color=contaminant)) + geom_point() +
  xlab("Prevalence (Negative Controls)") + ylab("Prevalence (True Samples)")
```
##remove contaminants
```{r Remove Contaminants}
physeq <- prune_taxa(!contamdf.prev$contaminant, physeq)
physeq
```

##Remove the sequence itself and replace with ASV
```{r}
dna <- Biostrings::DNAStringSet(taxa_names(physeq))
names(dna) <- taxa_names(physeq)
physeq <- merge_phyloseq(physeq, dna)
taxa_names(physeq) <- paste0("ASV", seq(ntaxa(physeq)))
physeq
```

##remove mitochondria and chloroplast matches. 
```{r Remove Mito and Chloro}
physeq <- physeq %>% subset_taxa( Family!= "Mitochondria" | is.na(Family) & Order!="Chloroplast" | is.na(Order) ) 
physeq
```

##remove all non bacterial sequences
```{r Remove Non-Bact}
physeq<-rm_nonbac(physeq)
physeq
```

#object representing time0
```{r}
physeq0 = subset_samples(physeq, time != "1" & time != "2" & time != "3" 
                         & time != "4" & time != "5")

physeq0
```

## Save physeq objects for easy loading
```{r}
save(physeq, file="../RData/physeq.RData")
save(physeq0, file="../RData/physeq0.RData")
```

##Alpha Diversity based on flock
```{r warning=FALSE, message=FALSE}
p=plot_richness(physeq,x="flock", measures=c("Observed","Simpson", "Shannon"))
flockalpha=p + geom_boxplot(data = p$data, aes(x = flock, y = value), alpha = 0.1) + theme(axis.title = element_text(face="bold"))
flockalpha
```

##Alpha Diversity based on sex with stats
```{r warning=FALSE, message=FALSE}
p=plot_richness(physeq,x="sex", measures=c("Observed","Simpson", "Shannon"))
BAR <- p + geom_boxplot(data = p$data, aes(x = sex, y = value, color = NULL), alpha = 0.1) + theme(axis.title = element_text(face="bold"))

bar <- BAR + stat_compare_means(aes(label = ifelse(..p.signif.. < 0.05, ..p.signif.., "")), method = "wilcox.test", label.x = 1.3, label.y = Inf)

barsex<-bar+ stat_compare_means(aes(label = ..p.signif..), method = "wilcox.test",  label.x = 1.5)

barsex
```

##Alpha Diversity based on population
```{r warning=FALSE, message=FALSE}
p=plot_richness(physeq,x="population", measures=c("Observed","Simpson", "Shannon"))
BAR <- p + geom_boxplot(data = p$data, aes(x = population, y = value, color = NULL), alpha = 0.1) + theme(axis.title = element_text(face="bold"))

bar <- BAR + stat_compare_means(aes(label = ifelse(..p.signif.. < 0.05, ..p.signif.., "")), method = "wilcox.test", label.x = 1.3, label.y = Inf)

barpop<-bar+ stat_compare_means(aes(label = ..p.signif..), method = "wilcox.test",  label.x = 1.5)

barpop
```

##Alpha Diversity based on flock t=0
```{r warning=FALSE, message=FALSE}
p0=plot_richness(physeq0,x="flock", measures=c("Observed","Simpson", "Shannon"))

flockalpha0=p0 + geom_boxplot(data = p0$data, aes(x = flock, y = value), alpha = 0.1) + xlab("Flock") + theme(axis.title = element_text(face = "bold"), 
          axis.title.x = element_text(face="bold"), plot.title = element_text(size = 10, face = "bold", hjust = .5))

flockalpha0
```

##Alpha Diversity based on sex with stats t=0
```{r warning=FALSE, message=FALSE}
p0=plot_richness(physeq0,x="sex", measures=c("Observed","Simpson", "Shannon"))

sexalpha0=p0 + geom_boxplot(data = p0$data, aes(x = sex, y = value), alpha = 0.1) + xlab("Sex") + theme(axis.title = element_text(face = "bold"), 
          axis.title.x = element_text(face="bold"), plot.title = element_text(size = 10, face = "bold", hjust = .5))

print(sexalpha0)
```

##Alpha Diversity based on population t=0
```{r warning=FALSE, message=FALSE}
p0=plot_richness(physeq0,x="population", measures=c("Observed","Simpson", "Shannon"))

popalpha0=p0 + geom_boxplot(data = p0$data, aes(x = population, y = value), alpha = 0.1) + xlab("Population") + theme(axis.title = element_text(face = "bold"), 
          axis.title.x = element_text(face="bold"), plot.title = element_text(size = 10, face = "bold", hjust = .5))

print(popalpha0)
```

##Alpha Diversity based on Time Point
```{r warning=FALSE, message=FALSE}
x <- seq(0,5,by=1)
p=plot_richness(physeq,x="timepoint", measures=c("Observed","Simpson", "Shannon"))
timepointalpha=p + geom_boxplot(data = p$data, aes(x = timepoint, y = value, color = NULL), alpha = 0.1) +theme(axis.title = element_text(face="bold"))
timepointalpha
```

##grid alpha div time 0
```{r}
alpha0=ggarrange(flockalpha0,                                                 
          ggarrange(sexalpha0, popalpha0, ncol = 2, labels = c("B", "C")), 
          nrow = 2, 
          labels = "A"                                        
          ) 

alpha0
```

##export tiff with 300dpi
```{r}
ggsave(
  filename="../figures/Figure01AlphaDiv0.tiff",
  plot = alpha0,
  width = 200,
  height = 200,
  units = c("mm"),
  dpi = 300,
)
```

##Export alpha diveristy
```{r}
# Estimate richness
alphadiv <- estimate_richness(physeq, measures = c("Observed", "Shannon", "Simpson"))

# Create a new column with the original sample names
alphadiv$sample <- gsub("^X", "", rownames(alphadiv))

# Write the modified data frame to a CSV file
write.csv(alphadiv, "../alphasheets/alpha_div.csv", row.names = FALSE)

```

##Merge metadata with alpha_div. Make sure sample column has header "sample" in both files
```{r}
# Read the CSV files
metadata <- read.csv("../metadata.csv")
alpha_div <- read.csv("../alphasheets/alpha_div.csv")

# Merge the dataframes by the 'sample' column
mergedalpha <- merge(metadata, alpha_div, by = "sample")

# Write the merged dataframe to a new CSV file
write.csv(mergedalpha, "../alphasheets/mergedalpha.csv", row.names = FALSE)
```

##hist
```{r}
par(mfrow = c(1, 3))
hist(mergedalpha$Shannon, main="Shannon diversity", xlab="", breaks=10)
hist(mergedalpha$Simpson, main="Simpson diversity", xlab="", breaks=10)
hist(mergedalpha$Observed, main="Observed ASVs", xlab="", breaks=10)
```

##Test for normality Shannon
```{r}
shapiro.test(mergedalpha$Shannon)
shapiro.test(mergedalpha$Simpson)
shapiro.test(mergedalpha$Observed)
```

## mixed effect model for population accounting for repeated sampling of birds
```{r}
# Fit a mixed-effects model for the Observed metric
model_observed <- lmer(Observed ~ population + (1 | bird), data = mergedalpha)

# Fit a mixed-effects model for the Shannon metric
model_shannon <- lmer(Shannon ~ population + (1 | bird), data = mergedalpha)

# Fit a mixed-effects model for the Simpson metric
model_simpson <- lmer(Simpson ~ population + (1 | bird), data = mergedalpha)

# Summary of the models
summary(model_observed)
summary(model_shannon)
summary(model_simpson)
```

## mixed effect model for sex accounting for repeated sampling of birds
```{r}
# Fit a mixed-effects model for the Observed metric
model_observed <- lmer(Observed ~ sex + (1 | bird), data = mergedalpha)

# Fit a mixed-effects model for the Shannon metric
model_shannon <- lmer(Shannon ~ sex + (1 | bird), data = mergedalpha)

# Fit a mixed-effects model for the Simpson metric
model_simpson <- lmer(Simpson ~ sex + (1 | bird), data = mergedalpha)

# Summary of the models
summary(model_observed)
summary(model_shannon)
summary(model_simpson)
```

## mixed effect model for flock accounting for repeated sampling of birds
```{r}
# Fit a mixed-effects model for the Observed metric
model_observed <- lmer(Observed ~ flock + (1 | bird), data = mergedalpha)

# Fit a mixed-effects model for the Shannon metric
model_shannon <- lmer(Shannon ~ flock + (1 | bird), data = mergedalpha)

# Fit a mixed-effects model for the Simpson metric
model_simpson <- lmer(Simpson ~ flock + (1 | bird), data = mergedalpha)

# Summary of the models
summary(model_observed)
summary(model_shannon)
summary(model_simpson)
```


##timepoint Friedman (repeated measures non parametric)
```{r warning=FALSE}
#Observed ASVs
friedman.test(y=mergedalpha$Observed, groups=mergedalpha$time, blocks=mergedalpha$bird)
pairwise.wilcox.test(mergedalpha$Observed, mergedalpha$time, p.adjust.method="holm")

#Shannon
friedman.test(y=mergedalpha$Shannon, groups=mergedalpha$time, blocks=mergedalpha$bird)
pairwise.wilcox.test(mergedalpha$Shannon, mergedalpha$time, p.adjust.method="holm")

#Simpson
friedman.test(y=mergedalpha$Simpson, groups=mergedalpha$time, blocks=mergedalpha$bird)
pairwise.wilcox.test(mergedalpha$Simpson, mergedalpha$time, p.adjust.method="holm")
```

##Alpha Diversity Time Point Observed
```{r}
observed<-ggscatter(mergedalpha, x = "time", y = "Observed", 
          add = "reg.line", conf.int = TRUE, cor.coef = TRUE, cor.coeff.args = list(method = "pearson", label.x = 3.2, label.sep = "\n"),
          xlab = "Time", ylab = "Observed ASVs")
observed
```

##Alpha Diversity Time Point Shannon
```{r}
shannon<-ggscatter(mergedalpha, x = "time", y = "Shannon", 
          add = "reg.line", conf.int = TRUE, cor.coef = TRUE, cor.coeff.args = list(method = "pearson", label.x = 2, label.y=4, label.sep = "\n"),
          xlab = "Time", ylab = "Shannon Diversity Index")
shannon
```

##Alpha Diversity Time Point Simpson
```{r}
simpson<-ggscatter(mergedalpha, x = "time", y = "Simpson", 
          add = "reg.line", conf.int = TRUE, cor.coef = TRUE, cor.coeff.args = list(method = "pearson", label.x = 3.2, label.y= .96, label.sep = "\n"),
          xlab = "Time", ylab = "Simpsons Index")
simpson
```

##grid
```{r}
timepoint=ggarrange(observed, shannon, simpson, 
          labels = c("A", "B", "C"),
          ncol = 3, nrow = 1)

timepoint
```

##export tiff with 300dpi
```{r}
ggsave(
  filename="../figures/Figure02Timepoint.tiff",
  plot = timepoint,
  width = 250,
  height = 100,
  units = c("mm"),
  dpi = 300,
)
```

##Remove taxa with relative abundance <0.005%
```{r}
minTotRelAbun = .00005
x = taxa_sums(physeq)
keepTaxa = (x / sum(x)) > minTotRelAbun
physeqprune = prune_taxa(keepTaxa, physeq)
physeqprune
```

##Remove taxa with relative abundance <0.005%
```{r}
minTotRelAbun = .00005
x = taxa_sums(physeq0)
keepTaxa = (x / sum(x)) > minTotRelAbun
physeqprune0 = prune_taxa(keepTaxa, physeq0)
physeqprune0
```
## Save physeq objects for easy loading
```{r}
save(physeqprune, file="../RData/physeqprune.RData")
save(physeqprune0, file="../RData/physeqprune0.RData")
```

##Determine prevelance of ASVs across all samples
```{r}
prevalence=prevalence(
  physeqprune,
  detection = 0,
  sort = FALSE,
  count = FALSE,
  include.lowest = FALSE
)

write.csv(prevalence,"../sheets/prevalence.csv")

# Read the CSV file
prevalence_data <- read.csv("../sheets/prevalence.csv")

# Create bins for every 0.05 interval
# Define breaks from 0 to 1 by 0.05
breaks <- seq(0, 1, by = 0.05)

# Cut the prevalence data into these bins
prevalence_data$bins <- cut(prevalence_data$x, breaks, include.lowest = TRUE)

# Count the number of ASVs in each bin
asv_counts <- table(prevalence_data$bins)

# Print the results
print(asv_counts)
```

##Create graph of ASV prevalence (data from prevelance.csv)
```{r}
data<-data.frame(
  NumberofASVs=c(2460,642,234,121,87,34,37,22,14,19,16,6,10,7,8,5,0,3,4,1),   
  SamplePercent=c(5,10,15,20,25,30,35,40,45,50,55,60,65,70,75,80,85,90,95,100))

asvprevalence=ggplot(data, aes(y=NumberofASVs, x=SamplePercent)) + 
  geom_bar(stat = "identity")

asvprevalence
```

##Bray Curtis Calculation
```{r message=FALSE}
dist = phyloseq::distance(physeqprune, method="bray", weighted=TRUE)
ordination = ordinate(physeqprune, method="NMDS", distance=dist)
```

##Bray Curtis Calculation time zero
```{r message=FALSE}
dist0 = phyloseq::distance(physeqprune0, method="bray", weighted=TRUE)
ordination0 = ordinate(physeqprune0, method="NMDS", distance=dist0)
```

##dist
```{r}
beta_boxplot <- function(physeqprune, method = "bray", group) {
  # Load required libraries
  library(phyloseq)
  library(ggplot2)

  # Convert 'timepoint' column to factor with correct levels
  physeqprune@sam_data$timepoint <- factor(physeqprune@sam_data$timepoint)

  # Identify the correspondence: group and samples
  group2samp <- list()
  group_list <- phyloseq::sample_data(physeqprune)[[group]]
  for (groups in levels(group_list)) {
    target_group <- which(group_list == groups)
    group2samp[[groups]] <- phyloseq::sample_names(physeqprune)[target_group]
  }

  # Calculate beta-diversity
  beta_div_dist <- phyloseq::distance(physeq = physeq, method = method)
  beta_div_dist <- as(beta_div_dist, "matrix")

  # Create a list of data frames for each group
  group_dfs <- lapply(names(group2samp), function(group_name) {
    group_samples <- group2samp[[group_name]]
    beta_div_values <- beta_div_dist[group_samples, group_samples]
    beta_div_values <- beta_div_values[lower.tri(beta_div_values)]

    data.frame(
      sample_pair = combn(group_samples, 2, paste, collapse = "-"),
      group = group_name,
      beta_div_method = method,
      beta_div_value = beta_div_values
    )
  })

  # Combine data frames into a single data frame
  dist_df <- do.call(rbind, group_dfs)

  # Convert the 'group' column to a factor with correct levels
  dist_df$group <- factor(dist_df$group, levels = levels(group_list))

  # Check if 'beta_div_value' contains any NA values
  if (any(is.na(dist_df$beta_div_value))) {
    stop("The 'beta_div_value' column contains NA values. There may be an issue with calculating beta-diversity.")
  }

  # Create a ggplot2 boxplot
  plot_boxplot <- ggplot(data = dist_df, aes(x = group, y = beta_div_value, color = group)) + 
    geom_boxplot(outlier.shape = NA, fill = "transparent", color = "black") +
    geom_jitter() + 
    theme_bw() + 
    xlab("Group") + ylab("Beta Diversity Value") +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    labs(x = "Time Point", y = "Bray-Curtis Dissimilarity")


  # Save df and boxplot into a list 
  list_Out <- list("data" = dist_df, "plot" = plot_boxplot) 

  return(list_Out)
}

# Test function 
beta_boxplot_result <- beta_boxplot(physeq = physeqprune, method = "bray", group = "timepoint")

## Data
beta_boxplot_result$data

## Plot
betabox <- beta_boxplot_result$plot

betabox
```
##export tiff with 300dpi
```{r}
ggsave(
  filename="../figures/300dpibetaboxdiv.tiff",
  plot = betabox,
  width = 240,
  height = 180,
  units = c("mm"),
  dpi = 300,
)
```

##Export beta diveristy
```{r}
write.csv(beta_boxplot_result$data, "../betasheets/beta_div.csv")
```

##Import betadiv after adding a column for time with numeric values instead of Time 0, Time 1, etc
```{r}
# Read the CSV file 
beta_div <- read.csv("../betasheets/beta_div.csv")

# Create a new 'time' column by removing the word 'Time' and converting to numeric
beta_div$time <- as.numeric(gsub("Time ", "", beta_div$group))

# Write the modified dataframe back to a CSV file
write.csv(beta_div, "../betasheets/beta_div_modified.csv", row.names = FALSE)

betadiv <- read.csv("../betasheets/beta_div_modified.csv")
```

##beta Diversity Time Point Observed
```{r}
braycurtistime<-ggscatter(betadiv, x = "time", y = "beta_div_value", 
          add = "reg.line", conf.int = TRUE, cor.coef = TRUE, cor.coeff.args = list(method = "pearson", label.x = 3, label.y = .6, label.sep = "\n"),
          xlab = "Time", ylab = "Bray-Curtis Dissimilarity")
braycurtistime
```

##export tiff with 300dpi
```{r}
ggsave(
  filename="../figures/300dpibetacorrdiv.tiff",
  plot = braycurtistime,
  width = 240,
  height = 180,
  units = c("mm"),
  dpi = 300,
)
```

##Bray Curtis Sex Plot
```{r}
braysex=plot_ordination(physeqprune, ordination, color="sex") + 
  theme_classic() +
  theme(strip.background = element_blank()) + stat_ellipse(aes(group=sex))
braysex
```

##Bray Curtis Sex Plot time 0
```{r}
braysex0=plot_ordination(physeq0, ordination0, color="sex") + 
  theme_classic() +
  theme(strip.background = element_blank()) + stat_ellipse(aes(group=sex))
braysex0
```

##Bray Curtis Sex Stats
```{r}
adonis2(dist ~ sample_data(physeqprune)$sex, strata = sample_data(physeqprune)$bird)
ps.disper<-betadisper(dist, sample_data(physeqprune)$sex)
permutest(ps.disper, pair=TRUE)
```

##Bray Curtis Sex Stats time 0
```{r}
adonis2(dist0 ~ sample_data(physeqprune0)$sex)
ps.disper0<-betadisper(dist0, sample_data(physeqprune0)$sex)
permutest(ps.disper0, pair=TRUE)
```

##Bray Curtis Population Plot
```{r}
braypop=plot_ordination(physeqprune, ordination, color="population") + theme_classic() +
  theme(strip.background = element_blank()) +  stat_ellipse(aes(group=population))
braypop
```

##Bray Curtis Population Plot time 0
```{r}
braypop0=plot_ordination(physeqprune0, ordination0, color="population") + theme_classic() +
  theme(strip.background = element_blank()) +  stat_ellipse(aes(group=population))
braypop0
```

##Bray Curtis Pop Stats
```{r}
adonis2(dist ~ sample_data(physeqprune)$population, strata = sample_data(physeqprune)$bird)
ps.disper<-betadisper(dist, sample_data(physeqprune)$population)
permutest(ps.disper, pair=TRUE)
```

##Bray Curtis Pop Stats time 0
```{r}
adonis2(dist0 ~ sample_data(physeqprune0)$population)
ps.disper0<-betadisper(dist0, sample_data(physeq0)$population)
permutest(ps.disper0, pair=TRUE)
```

##Bray Curtis Timepoint Plot
```{r}
braytimepoint=plot_ordination(physeq, ordination, color="timepoint") + 
  theme_classic() +
  theme(strip.background = element_blank()) + stat_ellipse(aes(group=timepoint))
braytimepoint
```

##Bray Curtis Time Point
```{r}
adonis2(dist ~ sample_data(physeqprune)$timepoint, permutations = 9999, strata = sample_data(physeqprune)$bird)
ps.disper<-betadisper(dist, sample_data(physeqprune)$timepoint)
permutest(ps.disper, pair=TRUE)
```

##Bray Curtis NMDS Flock Plot
```{r}
brayflock=plot_ordination(physeqprune, ordination, color="flock") + 
  theme_classic() +
  theme(strip.background = element_blank())+ stat_ellipse(aes(group=flock))
brayflock
```

##Bray Curtis NMDS Flock Plot time 0
```{r}
brayflock0=plot_ordination(physeqprune0, ordination0, color="flock") + 
  theme_classic() +
  theme(strip.background = element_blank())+ stat_ellipse(aes(group=flock))
brayflock0
```

##Bray Curtis Flock Stats 
```{r}
adonis2(dist ~ sample_data(physeqprune)$flock)
ps.disper<-betadisper(dist, sample_data(physeq)$flock)
permutest(ps.disper, pair=TRUE)
```

##Bray Curtis Flock Stats time 0
```{r}
adonis2(dist0 ~ sample_data(physeqprune0)$flock)
ps.disper0<-betadisper(dist0, sample_data(physeqprune0)$flock)
permutest(ps.disper0, pair=TRUE)
```

## Create file to analyze how pairs change
```{r}
# Read the metadata and beta diversity files
metadata <- read.csv("../metadata.csv")
beta_div_modified <- read.csv("../betasheets/beta_div_modified.csv")

# Extract the valid pairs and create the sample_pair column
sample_pairs <- metadata %>%
  filter(!is.na(pair)) %>%
  select(sample, pair, time) %>%
  distinct() %>%
  group_by(pair, time) %>%
  summarise(samples = list(sample), .groups = 'drop') %>%
  filter(lengths(samples) == 2) %>%
  rowwise() %>%
  mutate(
    sample1 = samples[1],
    sample2 = samples[2],
    sample_pair_1 = paste0(sample1, "-", sample2),
    sample_pair_2 = paste0(sample2, "-", sample1)
  ) %>%
  ungroup() %>%
  select(pair, time, sample_pair_1, sample_pair_2) %>%
  pivot_longer(cols = starts_with("sample_pair"), names_to = "key", values_to = "sample_pair") %>%
  select(pair, time, sample_pair)

# Check the resulting data frame
print(sample_pairs)

# Merge beta_div_modified with sample_pairs to retain only matching sample pairs and add the pair column
filtered_beta_div <- beta_div_modified %>%
  inner_join(sample_pairs, by = c("sample_pair" = "sample_pair"))

# Check the filtered and merged data frame
print(filtered_beta_div)

# Write the filtered and merged data to a new CSV file
write.csv(filtered_beta_div, "../betasheets/filtered_beta_div_modified_with_pairs.csv", row.names = FALSE)
```

##betadiv pair changes
```{r}
betadivpair<- read.csv("../betasheets/filtered_beta_div_modified_with_pairs.csv")

#test for normality
shapiro.test(betadivpair$beta_div_value)

#data are not normally distributed
#Friedman test - p=0.0739
friedman.test(y=betadivpair$beta_div_value, groups=betadivpair$time.x, blocks=betadivpair$pair)

#pairwise
pairwise.wilcox.test(betadivpair$beta_div_value, betadivpair$time.x, p.adjust.method="holm")

# Ensure the 'time' column is treated as a factor (if it's categorical)
betadivpair$time.x <- as.factor(betadivpair$time.x)

# Perform Levene's Test
result <- leveneTest(beta_div_value ~ time.x, data = betadivpair)
result
```

```{r}
##graph
betadivpair$group <- as.factor(betadivpair$group)
betapair<-ggscatter(betadivpair, x = "time.y", y = "beta_div_value", 
          add = "reg.line", conf.int = TRUE, cor.coef = TRUE, cor.coeff.args = list(method = "pearson", label.x = 3.1, label.sep = "\n"),
          xlab = "Time", ylab = "Bray-Curtis Distance")
betapair
```

##grid bray
```{r}
bray=ggarrange(ggarrange(braysex, brayflock, braypop, braytimepoint, 
          labels = c("A", "B", "C", "D"),
          ncol = 2, nrow=2), betabox, nrow=2, labels=c("","E"), heights = c(2.5,1.5))

bray
```

##export tiff with 300dpi
```{r}
ggsave(
  filename="../figures/Figure04BetaDiv.tiff",
  plot = bray,
  width = 250,
  height = 200,
  units = c("mm"),
  dpi = 300,
)
```



##Bar plots of Abundance per individual samples in timepoint (Phylum-Merge <1%)
```{r}
physeq2 = filter_taxa(physeq, function(x) mean(x) > 0.01, TRUE)
physeq3 = transform_sample_counts(physeq2, function(x) x / sum(x) )
glom<-psmelt(physeq3)
glom <- tax_glom(physeq3, taxrank = 'Phylum')
data<-psmelt(glom)
data$Phylum <- as.character(data$Phylum) 
data$Phylum[data$Abundance < 0.01] <- "< 1% abund."
medians <- ddply(data, ~Phylum, function(x) c(median=median(x$Abundance)))
remainder <- medians[medians$median <= 0.01,]$Phylum
data[data$Phylum %in% remainder,]$Phylum <- "< 1% abund."
data$Phylum[data$Abundance < 0.010] <- "< 1% abund."
spatial_plot <- ggplot(data=data, aes(x=bird, y=Abundance, fill=Phylum)) + 
  facet_wrap(~timepoint, scales = "free")
barplotphylum<-spatial_plot + geom_bar(aes(), stat="identity", position="fill") + 
  ggtitle("Phylum Abundance at Each Time Point") + 
  theme (axis.text.x = element_text(angle=90), 
         plot.title = element_text(size = 10, face = "bold", hjust = .5)) 
barplotphylum
```
##export tiff with 300dpi
```{r}
ggsave(
  filename="../figures/Figure00Barplot.tiff",
  plot = barplotphylum,
  width = 250,
  height = 150,
  units = c("mm"),
  dpi = 300,
)
```

##export otu table of phylum abundance
```{r}
# Extract abundance matrix from the phyloseq object
OTU1 = as(otu_table(glom), "matrix")
# transpose if necessary
if(taxa_are_rows(glom)){OTU1 <- t(OTU1)}
# Coerce to data.frame
OTUdf = as.data.frame(OTU1)
#export to csv
write.csv(OTUdf, "../sheets/phylumdf.csv", row.names = TRUE)
```

##Bar plots of Abundance per individual samples in timepoint (Order-Merge <10%)
```{r}
physeq2 = filter_taxa(physeq, function(x) mean(x) > 0.1, TRUE)
physeq3 = transform_sample_counts(physeq2, function(x) x / sum(x) )
glom<-psmelt(physeq3)
glom <- tax_glom(physeq3, taxrank = 'Order')
data<-psmelt(glom)
data$Order <- as.character(data$Order) 
data$Order[data$Abundance < 0.1] <- "< 10% abund."
medians <- ddply(data, ~Order, function(x) c(median=median(x$Abundance)))
remainder <- medians[medians$median <= 0.1,]$Order
data[data$Order %in% remainder,]$Order <- "< 10% abund."
data$Order[data$Abundance < 0.10] <- "< 10% abund."
spatial_plot <- ggplot(data=data, aes(x=bird, y=Abundance, fill=Order)) + 
  facet_wrap(~timepoint, scales = "free")
barplotorder<-spatial_plot + geom_bar(aes(), stat="identity", position="fill") + 
  ggtitle("Order Abundance at Each Time Point") + 
  theme (axis.text.x = element_text(angle=90), 
         plot.title = element_text(size = 10, face = "bold", hjust = .5)) 
barplotorder
```
##export tiff with 300dpi
```{r}
ggsave(
  filename="../figures/Figure00BarplotOrder.tiff",
  plot = barplotorder,
  width = 250,
  height = 150,
  units = c("mm"),
  dpi = 300,
)
```

#Differential species ID
```{r warning=FALSE}
# Extract abundance data (OTU table)
abundance_data <- as.data.frame(otu_table(physeqprune))

# Extract the grouping variable (time) from sample data
time_group <- sample_data(physeqprune)$time

# Ensure the grouping variable is a factor
time_group <- as.factor(time_group)

# Initialize a list to store SIMPER results
all_simper_results <- list()

# Get unique pairs of time points
time_pairs <- combn(unique(time_group), 2, simplify = FALSE)

# Run SIMPER analysis for each pairwise comparison
for (pair in time_pairs) {
  time1 <- pair[1]
  time2 <- pair[2]
  comparison_label <- paste(time1, time2, sep = "_")

  # Subset data for the pair
  subset_data <- subset_samples(physeqprune, time %in% pair)
  subset_abundance_data <- as.data.frame(otu_table(subset_data))
  subset_time_group <- sample_data(subset_data)$time

  # Run SIMPER analysis
  simper_result <- simper(subset_abundance_data, group = as.factor(subset_time_group), permutations = 100)
  simper_summary <- summary(simper_result)

  # Check if simper_summary has the comparison_label data frame
  if (!is.null(simper_summary[[comparison_label]])) {
    # Extract the data frame from the list and add comparison information
    simper_df <- as.data.frame(simper_summary[[comparison_label]])
    simper_df$species <- rownames(simper_df)
    rownames(simper_df) <- NULL
    simper_df$comparison <- comparison_label

    # Store results
    all_simper_results[[comparison_label]] <- simper_df
  }
}

# Combine all SIMPER results into a single data frame
combined_simper_df <- do.call(rbind, all_simper_results)

# Save combined SIMPER results to a CSV file
write.csv(combined_simper_df, "../differentialexpression/simper_results.csv", row.names = FALSE)

# Extract OTU names and their contributions
otu_contributions <- combined_simper_df %>%
  select(species, average, sd, ratio, ava, avb, cumsum, comparison) %>%
  arrange(species)

# Initialize a data frame to store Mann-Whitney U Test results
results <- data.frame()

# Perform Mann-Whitney U Test on OTU contributions between groups
for (otu in unique(otu_contributions$species)) {
  otu_data <- subset(otu_contributions, species == otu)

  for (comparison in unique(otu_data$comparison)) {
    comparison_data <- subset(otu_data, comparison == comparison)

    # Perform Mann-Whitney U Test
    test_result <- wilcox.test(comparison_data$ava, comparison_data$avb)

    # Store results
    results <- rbind(results, data.frame(OTU = otu, comparison = comparison, p.value = test_result$p.value))
  }
}

# Adjust p-values for multiple testing (optional)
results$adj.p.value <- p.adjust(results$p.value, method = "BH")

# Save Mann-Whitney results to a CSV file
write.csv(results, "../differentialexpression/mann_whitney_results.csv", row.names = FALSE)

# View significant results
significant_results <- subset(results, adj.p.value < 0.05)

# Save significant results to a CSV file
write.csv(significant_results, "../differentialexpression/significant_results.csv", row.names = FALSE)

# Print significant results
print(significant_results)
```